\chapter{How Cricket Works}
\label{chap: How Cricket Works}

I recommend watching this short two minute video which explains the game succinctly. Linked here: \cite{global_news_cricket_2019}

\subsubsection{Transcript from the video}

Cricket is played on an oval shaped field. In the middle of the field is a rectangle of hard turf known as the pitch. At each end of the pitch are wickets, made up of three wooden stumps to support two bails. The game involves two teams with 11 players making up each team. The aim of the game is to score more runs than the opposition. The umpires judge fair and unfair play. A match begins with one team going out on the field and the other comes into bat. The fielding team has a bowler, who uses different techniques to bowl the ball to the batter. The batter tries to hit the ball and score runs by hitting the ball and running between the wickets; making it to the other end of the pitch before the fielders can hit the wickets with the ball. Each time the batter runs a full length of the pitch, he scores 1 run for his team. Hitting the ball to the boundary, along the ground is 4 runs. Hitting the ball over the boundary without touching the ground is 6 runs - the highest scoring shot. In the meantime, the fielding team tries to get the batter out as quickly as possible in a number of ways. Knocking the bails off the stumps with the ball when bowling (bowled out); catch a batters shot, before it touches the ground (caught out); hitting the batter's leg if their leg was deemed by the umpire to be blocking the ball's path to the stumps (leg before wicket); or hitting the wickets before the batters can run to the other side of the pitch (run out). Each set of six legal balls bowled, constitutes an over. Our focus is on a shortened form of cricket, Twenty20. In this form of the game, each innings lasts for a maximum of 20 overs (120 balls). Once all the batters are out, or if they run out of overs, the innings comes to a close and the fielding team comes into bat, with the goal of chasing down the first innings score, within the allotted 20 overs.